---
title: "STAT632_final_project_code_appendix"
author: "Shonushka S, Megan M, Bianca C"
date: "May 7, 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Code Appendix for STAT 632 Final Project

**Load the Necessary Libraries & Data Set**

```{r data_import, include = TRUE}
library(ggplot2)
library(dplyr)
library(Stat2Data)
library(cowplot)
library(pROC)
library(car)
data(BirdNest)
#head(BirdNest)
```

**Cleaning the data set**

```{r cleaning, results = 'hide', echo = TRUE}
# removed observation without entries
nests <- na.omit(BirdNest)

# change Closed. to factor
nests$clo<-as.factor(nests$Closed.)
```

**For data summaries**

```{r stat_summary, results = 'hide', echo = TRUE}
# summary statistics for numerical variables
nests %>%
  summarise(meanLength = round(mean(Length), 2),
            sdLength = round(sd(Length), 2),
            meanEggs = round(mean(No.eggs), 2),
            sdEggs = round(sd(No.eggs), 2),
            meanIncu = round(mean(Incubate), 2),
            sdIncu = round(sd(Incubate), 2),
            meanNest = round(mean(Nestling), 2),
            sdNest = round(sd(Nestling), 2),
            meanCare = round(mean(Totcare), 2),
            sdCare = round(sd(Totcare), 2))

# count number of observations in Color and Closed
nests %>% count(Color == 0)
nests %>% count(Closed. == 0)
```

**For histogram plots**

```{r hist_plots, results = FALSE, show.fig = 'hide'}

#library(cowplot)
bird_colors <- c("#fda630", "#f86734", "#e867ad", "#9b35dd",
                 "#7968f0", "#3e45a2", "#161b61", "#af2405", "#d1d1d1")

# NUMBER OF EGGS
p1 <- ggplot(nests, aes(x = No.eggs)) +
  geom_histogram(binwidth = 1, fill = "#fda630", color = "white", alpha = 0.8) +
  labs(x = "Number of Eggs", y = "Count") +
  theme_classic() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))


# BODY LENGTH
p2 <- ggplot(nests, aes(x = Length)) +
  geom_histogram(binwidth = 2.5, fill = "#f86734", color = "white", alpha = 0.6) +
  labs(x = "Mean Body Length", y = "Count") +
  theme_classic() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))

# INCUABTE TIME
p3<- ggplot(nests, aes(x = Incubate)) +
  geom_histogram(binwidth = 1, fill = "#e867ad", color = "white", alpha = 0.4) +
  labs(x = "Mean Incubation (days)", y = "Count") +
  theme_classic() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))

# NESTIING TIME
p4 <- ggplot(nests, aes(x = Nestling)) +
  geom_histogram(binwidth = 1, fill = "#9b35dd", color = "white", alpha = 0.3) +
  labs(x = "Mean Nestling (days)", y = "Count") +
  theme_classic() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12))

# plot all the histograms
histos <- plot_grid(p1, p2, p3, p4)
ggdraw(add_sub(histos, "Figure 1: The variables length, 
               number of eggs, and incubation all have a normal 
               distribution with a right-skew.", size = 8))
```

**For barplots**

```{r bar_plots, results = 'hide', show.fig = 'hide'}
# LOCATION
loc_num <- nests %>% count(Location)
loc_count <- loc_num[,2]

b1 <- ggplot(nests, aes(x = Location, fill = "#af2405")) +
  geom_bar(alpha = 0.7) +
  theme_classic() +
  labs(x = "Nest Location", y = "Count") +
  annotate("text", x = 1:9, y = loc_count + 1, label = loc_count) +
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12),
        legend.position = "none")


# NEST TYPE
type_num <- nests %>% count(Nesttype)
type_count <- type_num[,2]

b2 <- ggplot(nests, aes(x = Nesttype, fill = clo)) +
  geom_bar(alpha = 0.8) +
  theme_classic() +
  labs(x = "Nest Type", y = "Count") +
  annotate("text", x = 1:7, y = type_count + 1.2, label = type_count) +
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12),
        legend.position = c(0.8,0.8)) +
  scale_fill_manual(values = c("#7968f0", "#161b61"), name = "Type",
                    labels = c("Open", "Closed"))


# plot the bar charts
bars <- plot_grid(b1, b2)
ggdraw(add_sub(bars, "Firgure 2: There is an under representation of certain nest 
               locations in this data set, and an over representation of cup type nests.
               \nThis uneven representation may affect the final model.", size = 8))
```

**Full Multiple Logistic Regression Model**

```{r full_mlr_mod}
# Note that neither clo has been coded as factor
glm1 <- glm(clo ~ Length + No.eggs + Color + Incubate + Nestling,
           data = nests, family = binomial)
summary(glm1)

AIC(glm1)
```

**Step-wise Selection**

```{r stepwise_selection2}
# stepwise selection of variables
fbm <- step(glm1, trace = FALSE)
summary(fmb)
```

**Final Multiple Logistic Regression Model**

```{r final_mod, echo = TRUE}
# define the reduced model
glm2 <- glm(clo ~ Length + Nestling + No.eggs, data=nests, family=binomial)
summary(glm2)

# compare the AIC for full and reduced model
AIC(glm1, glm2)
```

**Logistic Regression Plots**

```{r log_reg_plots, results = 'hide', show.fig = 'hide'}
# graph for Length and Closed
m1 <- ggplot(nests, aes(Length, Closed.)) + 
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              se=F, color="blue") +
  xlab("Length of bird's body") +
  ylab("probability of nest being closed") + 
  theme_bw() +
  annotate(geom = "point", x = c(20, 22.5, 15), 
           y = c(0.1367266, 0.1348924, 0.4359303),
           color = c('darkred', 'cyan2', 'darkorange'),
           size = 3, shape = 13)

# graph for No.eggs and Closed
m2 <- ggplot(nests, aes(No.eggs, Closed.)) + 
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              se=F, color="blue") +
  xlab("Number of eggs") +
  ylab("probability of nest being closed") + 
  theme_bw() +
  annotate(geom = "point", x = c(4.5, 4.5, 5.0), 
           y = c(0.1367266, 0.1348924, 0.4359303),
           color = c('darkred', 'cyan2', 'darkorange'),
           size = 3, shape = 13)

# graph for Nestling and Closed.
m3 <- ggplot(nests, aes(Nestling, Closed.)) + 
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              se=F, color="blue") +
  xlab("Nestling time") +
  ylab("Probability of nest being closed") + 
  theme_bw() +
  annotate(geom = "point", x = c(15, 16.5, 15.5), 
           y = c(0.1367266, 0.1348924, 0.4359303),
           color = c('darkred', 'cyan2', 'darkorange'),
           size = 3, shape = 13)

logits <- plot_grid(m1, m2, m3, nrow = 1)
ggdraw(add_sub(logits, "Figure 3: Individual logit plots for each variable included in the final model. 
               \nFrom the cross-validation exercise, three species that have closed nests are 
               \npredicted to have open nests by the training model, are plotted above. 
               \nRed - Myiarchus cinerascens; Cyan - Myiarchus tyrannulus; Orange - Passer domesticus",
               size = 8))
```

**Training and Test Sets for Validation**

```{r train_and_test}
# there are 83 observations in BirdNest
floor(0.7*nrow(nests))
# 58 is 70% of the observations

# set seed!
set.seed(504)

# randomly sample 70% of the observations
nest_train <- sample(1:83, 58)

# fit the model using train
glm_train <- glm(Closed. ~ Length + No.eggs + Nestling, data = nests,
                 subset = nest_train, family = binomial)

# Are the variabels still significant?
summary(glm_train)

# subset data frame for testing observations
nest_test <- nests[-nest_train,]

# make predictions for probabilities on nest_test
probs_test <- predict(glm_train, newdata = nest_test, type = "response")

# remember thre are 25 observationsin the test set
preds_test <- rep(0, 25)
preds_test[probs_test > 0.5] <- 1

# make the confusion matrix table
tb <- table(prediction = preds_test,
            actual = nest_test$Closed.)
addmargins(tb)

# ACCURACY - percent all nests correctly classified
(tb[1, 1] + tb[2,2]) / 25

# SENSITIVITY - closed nests correctly classified
tb[2,2] / 10

# SPECIFICITY - open nests correctly classified
tb[1,1] / 15

# Which three CLOSED nests were incorrectly classified as OPEN?
# first add preds_test column to nest_nest df
nest_test <- nest_test %>% mutate(predict = preds_test)
nest_test <- nest_test %>% mutate(probabilty = probs_test)
# filter for observations with incorrectly classified OPEN nests
nest_test %>% filter(Closed. == 1 & predict == 0)


# compare closed vs open nests by predictors
nests %>%
  group_by(Closed.) %>%
  summarise(meanLength = round(mean(Length), 2),
            sdLength = round(sd(Length), 2),
            meanEggs = round(mean(No.eggs), 2),
            sdEggs = round(sd(No.eggs), 2),
            meanNest = round(mean(Nestling), 2),
            sdNest = round(sd(Nestling), 2))
```

**ROC and AUC Calculations**

```{r ROC_AUC, results = 'hide', message = FALSE, out.width='80%', warning = FALSE}
# library(pROC)
roc_obj <- roc(nest_test$Closed., probs_test)
plot(1 - roc_obj$specificities, roc_obj$sensitivities, type="l",
xlab = "1 - Specificity", ylab = "Sensitivity")
# plot red point corresponding to 0.5 threshold:
abline(0, 1, lty=2) # 1-1 line
auc(roc_obj)
```

**Passerine Predictions**

```{r bird_predictions}
# marsh wren should have a closed nest
marshwren <- data.frame(Length = 11.95, No.eggs = 5, Nestling = 14)
predict(glm2, newdata = marshwren, type = "response")

# tree swallow should have a closed nest
treeswallow <- data.frame(Length = 13, No.eggs = 5.5, Nestling = 20)
predict(glm2, newdata = treeswallow, type = "response")

# buff-breasted flycatcher should have an open nest
bbflycatcher <- data.frame(Length = 12.25, No.eggs = 3.4, Nestling = 16)
predict(glm2, newdata = bbflycatcher, type = "response")
```
